Change in Mobility and Site Occupation during the Late Pleistocene in Korea

### 2019-11-25 analysis/paper/paper.Rmd
make site elevation plot work

The data analyzed in this paper are gathered from published excavation reports of 22 Paleolithic sites (33 cultural layers) in South Korea dating from 49k BP to 24k BP. After the first excavation at Jeongokri site in the late 1970s, more than 200 Paleolithic sites are discovered in Korea. Among them, we  chose 22 sites based on the existence of absolute dates that match the targeting timeline and officially published excavation report. We include sites dated before and after the transition period (40-35ka) to analyze the process of technological transition.

```{r cache = TRUE}
#I couldn't find the most recent map that you made. I assume that you don't share the code, but showed me the image so that I just copied it for SAA poster
#This is what I made first before you made the recent one

library(here)

@@

Assemblages
The research analyzes total 33 assemblages and they consist of cores, blade cores, blades, flakes, debris, hammers, choppers, plans, polyhedrals, side scrapers, end scrapers, notches, cleavers, stemmed points, awls, denticulates, burins, handaxes, blanks, knives, handadzes, peaks, flatters, flakers etc, 
Some artifacts are not included as certain tools due to their vague or broken shape presumably related to the characteristics of raw materials such as quartz, quartz vein, and quartzite. In that case, we named them as unknown, unfinished, or pebble. The stone artifacts from all of these sites are stored in museums located throughout South Korea, and have been briefly described in the excavation reports. We have visited most of these collections, collected PDFs of the excavation reports, and negotiated permission to study the materials. For example, the Hwadae-ri site has three cultural layers. The lowest horizon at the bottom, dated to 39,000±1400 BP by OSL, contains coarse flake tools made of vein quartz and quartzite. The middle cultural horizon has stemmed points made of porphyry, with quartz and quartzite dominating the assemblage. This layer dated to 31,200±900 BP by radiocarbon dating and 30,000±1,700 BP by OSL. The uppermost layer, dated to 22,000±100 BC by OSL, contains blades, scrapers, awls, and denticulates [@seong2009emergence]. We compare the assemblages from the two older layers at this and other sites to investigate the transition process.

@@

The artifact volumetric density is defined as the total number of artifacts per cubic meter of excavated sediment, and served as a proxy for accumulation ratio of artifacts [@clark2017lithics; @clark2019landscapes]. We expected to use this value to evaluate mobility and land-use practices of hunter-gatherers during the Late Pleistocene. We count the retouch frequency as the number of retouched tools divided by the total number of artifacts. This value represents general proficiency of lithic manufacturing, raw material availability, and subsistence strategies including mobility and duration of site occupation based on the concept that tool reductioning is an economical tactic to generate sharp, usable edges while minimizing the cost of transporting tools or bulky raw materials by reducing the total load weight [@Kuhn_1990; @clark2019landscapes]. In other words, producing and carrying highly retouched tools will increase portability and efficiency by decrease in the size of both individual tools and the assemblages so that hunter-gatherers could move often or further, or both [@kuhn1991unpacking; @kuhn1994formal; @kuhn2004upper; @andrefsky1994raw; @davies2018modeling]. For example, highly retouched and finely shaped implements appeared in Australia during the mid-Holocene to minimize risk associated with environmental change and high mobility [@Hiscock_1994].
Combined the two proxies, the artifact volumetric density and the retouch frequency, we expect to determine whether or not the hunter-gatherers were residentially mobile or logistically organized. If the assemblage in a site has high density and less retouched, it is depicted as expedient assemblage that represents “base camps” or “residences”, while less density with high retouched one shows curated assemblage and indicates either residential mobility or certain task groups from main residential site [@Riel-Salvatore_Barton_2004; @clark2017lithics].

@@

Elevation of occupied sites
The result of the elevation change of 33 cultural layers from 22 sites corresponding to the targeting period shows that there is no remarkable change can be seen in site elevation during the transition (Fig.1: with no sp site distinction). We applied the existence of the stemmed points to the elevation to understand the benefit of having the points in the toolkit and the result indicates that the hunter-gathering groups who carried the technology could occupy higher altitude while the groups with no points stayed at the lower altitude (Fig. 2: distinction between sites with no stemmed points and with stemmed points).

```{r}
#site elevation
mydata_ages <- 
  mydata %>% 
  separate(C14.BP., into = c('age', 'error'),
          sep = "±") %>% 
  mutate(age_ka = parse_number(age) / 1000,
         error = parse_number(error)) %>% 
  mutate(has_sp = ifelse(!is.na(SP.), "yes", "no"))

ggplot(mydata_ages, 
       aes(x = age_ka,
           y = altitude.m._of_main_layer,
       color = as.factor(has_sp))) +
  geom_point(size = 3) +
  xlab("Age of occupation (ka)") +
  ylab("Elevation above sea level (m)") +
  geom_smooth(se = FALSE) +
  scale_colour_discrete(name = "Contains\nstemmed\npoints?") +
  theme_minimal(base_size = 16) 
```

Artifact density and Composition of toolkit
During the transitional period, the artifact volumetric density was relatively small but it becomes higher around 30 ka (Fig. 3: the transition of artifact volumetric density). When it comes to the composition of the toolkit, it is hardly recognized any pattern of changes over time (Fig. 4: the composition of toolkit). Both results with no information about raw material availability near the 22 sites, we assume that the artifact density might relate to the duration of site occupation or group size rather than portability and efficiency of tools. In other words, the small artifact density might indicate shorter duration or smaller group size. Likewise, the composition of toolkit might represent the characteristics or purpose of each site, rather than showing general pattern of chronological change.

@@

Summed probability distribution
To understand the population change during the appearance of stemmed points, we analyzed the summed probability distributions of 171 radiocarbon dates from 33 cultural layers and it indicates that there is a higher density of radiocarbon dates from around 34 ka. We could cautiously assume that the population flux might relate to the technological innovation.

@@

Raw materials
During the transition period, the use of shale and rhyolite greatly increased. These two raw materials are suitable to make elongate blades and stemmed points. It is considered that he hunter-gatherer started to use raw materials selectively depending on the type of tool, especially the newly produced stemmed points. Raw material availability is a crucial factor in determining the trip and direction for hunter-gatherers [@davies2018modeling]. Therefore the adoption of new raw material along with the emergence of new type of tools demonstrates a significant innovation in not just technology but also cognitive system.

@@

1. The new technology and occupation in high elevation
The new technology appeared around 40-35ka in the Korean Peninsula includes stemmed points and following blade industry. The stemmed points brought new paradigms in stone artifact assemblages. For example, projectile points are only a part of composit artifacts and this specific characteristics could contribute to increase efficiency of maintaining the toolkit by easily replacing the damaged part [@kuhn2015artifacts; @cardillo2010some]. Combining with other materials increases utility or functionality of the tool by adding extra weight and providing a shaft, which makes it earlier to grip. Selective use of raw material might impact on hunter-gathers’ mobility and occupation pattern depending on availability of the specific raw materials such as shale and rhyolite. In addition, this research shows that carrying the points enabled hunter-gatherers to occupy higher elevation compared to the sites with no stemmed points. Therefore, we can assume that the technology would play an important role to survive in more risky environments.

2. Population change
Interpreted in HBE, we expected there might be close connection between environmental change and the appearance of new technology based on the prediction that the new invention was the result of risk. The research suggests that the environmental pressure could be related to population change. We cannot simply predicate that the increased population around 35ka was due to dispersal of Homo Sapiens since there is no reliable human remains in South Korea. However, the migration of Homo sapiens was not impossible scenario at the same time. In fact, one model explains that the emergence of stemmed points and blade industry came from outside of the Korean peninsula and the new technology was the result of the continuous influx of modern human populations from Siberia while the earlier coarse flake trandition was come from southern China [@Bae_2010]. Since this migration model barely explains about preexisted stone artifact assemblages and people in the Korean Peninsula, it is hard to accept his argument without knowing about the native people. In short, we assume that the stemmed points appeared related to the increased population and it can be related to the modern human dispersal or indigenously increased occupants.

3. Expedient vs curated
To monitor mobility strategies and site occupation pattern, we examined the artifact density and retouch frequency, based on the premise that highly mobile foragers would have light weighted, highly retouched, replaceable, almost completely made tools and leave small amount of artifacts when they move to the new location (curated). One the other hand, less mobile foragers who tend to stay in one place yield more debitage and debris, produce less efficient tools since they can easily get supplied (expedient) [@clark2019landscapes; @kuhn1994formal; @kuhn2014mousterian; @meignen2006middle]. Therefore, the curated technology indicates sites with special or highly mobile groups with stone artifacts that are thoroughly designed and based on a high level of investment of time and effort in artifact manufacture and maintenance. The expedient technology represents base camp or residence with artifacts that tend to be manufactured, used, and abandoned in the short period of time [@vaquero2018searching;@binford1979organization; @torrence1989time]. There appear to be more variety in the type of sites (or group) during the transition period, both expedient or curated. It became more expedient at the end of the transition, which means foragers became less mobile and tended to stay in the base camp. Combined with the result of population dynamics, we can assume that the increased population made each group bigger and let them stay in one place.

>>>>

The data analyzed in this paper are gathered from published excavation reports of 22 Paleolithic sites (33 cultural layers) in South Korea dating from 49k BP to 24k BP. After the first excavation at Jeongokri site in the late 1970s, more than 200 Paleolithic sites are discovered in Korea. Among them, we  chose 22 sites based on the existence of absolute dates that span our period of interest. We include sites dated before and after the transition period (40-35 ka) to analyze the process of technological transition.

```{r cache = TRUE, fig.cap="Map of sites discussed in the text."}
library(here)

@@

Assemblages

The research analyzes total 33 assemblages [need to explain what is the difference between 'assemblage' and 'site'] and they consist of cores, blade cores, blades, flakes, debris, hammers, choppers, plans, polyhedrals, side scrapers, end scrapers, notches, cleavers, stemmed points, awls, denticulates, burins, handaxes, blanks, knives, handadzes, peaks, flatters, flakers etc,

Some artifacts are not included as certain tools due to their vague or broken shape presumably related to the characteristics of raw materials such as quartz, quartz vein, and quartzite. In that case, we named them as unknown, unfinished, or pebble. The stone artifacts from all of these sites are stored in museums located throughout South Korea, and have been briefly described in the excavation reports. We have visited most of these collections, collected PDFs of the excavation reports, and negotiated permission to study the materials. For example, the Hwadae-ri site has three cultural layers. The lowest horizon at the bottom, dated to 39,000±1400 BP by OSL, contains coarse flake tools made of vein quartz and quartzite. The middle cultural horizon has stemmed points made of porphyry, with quartz and quartzite dominating the assemblage. This layer dated to 31,200±900 BP by radiocarbon dating and 30,000±1,700 BP by OSL. The uppermost layer, dated to 22,000±100 BC by OSL, contains blades, scrapers, awls, and denticulates [@seong2009emergence]. We compare the assemblages from the two older layers at this and other sites to investigate the transition process.

@@

The artifact volumetric density is defined as the total number of artifacts per cubic meter of excavated sediment, and served as a proxy for accumulation ratio of artifacts [@clark2017lithics; @clark2019landscapes]. We expected to use this value to evaluate mobility and land-use practices of hunter-gatherers during the Late Pleistocene. We count the retouch frequency as the number of retouched tools divided by the total number of artifacts. This value represents general proficiency of lithic manufacturing, raw material availability, and subsistence strategies including mobility and duration of site occupation based on the concept that tool reductioning is an economical tactic to generate sharp, usable edges while minimizing the cost of transporting tools or bulky raw materials by reducing the total load weight [@Kuhn_1990; @clark2019landscapes]. In other words, producing and carrying highly retouched tools will increase portability and efficiency by decrease in the size of both individual tools and the assemblages so that hunter-gatherers could move often or further, or both [@kuhn1991unpacking; @kuhn1994formal; @kuhn2004upper; @andrefsky1994raw; @davies2018modeling]. For example, highly retouched and finely shaped implements appeared in Australia during the mid-Holocene to minimize risk associated with environmental change and high mobility [@Hiscock_1994].

Combined the two proxies, the artifact volumetric density and the retouch frequency, we expect to determine whether or not the hunter-gatherers were residentially mobile or logistically organized. If the assemblage in a site has high density and less retouched, it is depicted as expedient assemblage that represents “base camps” or “residences”, while less density with high retouched one shows curated assemblage and indicates either residential mobility or certain task groups from main residential site [@Riel-Salvatore_Barton_2004; @clark2017lithics].

@@

Elevation of occupied sites

The result of the elevation change of 33 cultural layers from 22 sites corresponding to the targeting period shows that there is no remarkable change can be seen in site elevation during the transition (Fig.1: with no sp site distinction). We applied the existence of the stemmed points to the elevation to understand the benefit of having the points in the toolkit and the result indicates that the hunter-gathering groups who carried the technology could occupy higher altitude while the groups with no points stayed at the lower altitude (Fig. 2: distinction between sites with no stemmed points and with stemmed points).

```{r}
# site elevation

mydata <- korean_archaeological_sites

mydata_ages <- 
  mydata %>% 
  separate(`C14(BP)`, into = c('age', 'error'),
          sep = "±") %>% 
  mutate(age_ka = parse_number(age) / 1000,
         error = parse_number(error),
         `altitude(m)_of_main_layer` = parse_number(`altitude(m)_of_main_layer`)) %>% 
  mutate(has_sp = ifelse(!is.na(`SP#`), "yes", "no"))

ggplot(mydata_ages, 
       aes(x = age_ka,
           y = `altitude(m)_of_main_layer`,
       color = as.factor(has_sp))) +
  geom_point(size = 3) +
  xlab("Age of occupation (ka)") +
  ylab("Elevation above sea level (m)") +
  geom_smooth(se = FALSE) +
  scale_colour_discrete(name = "Contains\nstemmed\npoints?") +
  theme_minimal(base_size = 16) 
```

Artifact density and Composition of toolkit

During the transitional period, the artifact volumetric density was relatively small but it becomes higher around 30 ka (Fig. 3: the transition of artifact volumetric density). When it comes to the composition of the toolkit, it is hardly recognized any pattern of changes over time (Fig. 4: the composition of toolkit). Both results with no information about raw material availability near the 22 sites, we assume that the artifact density might relate to the duration of site occupation or group size rather than portability and efficiency of tools. In other words, the small artifact density might indicate shorter duration or smaller group size. Likewise, the composition of toolkit might represent the characteristics or purpose of each site, rather than showing general pattern of chronological change.

@@

Summed probability distribution

To understand the population change during the appearance of stemmed points, we analyzed the summed probability distributions of 171 radiocarbon dates from 33 cultural layers and it indicates that there is a higher density of radiocarbon dates from around 34 ka. We could cautiously assume that the population flux might relate to the technological innovation.

@@

Raw materials

During the transition period, the use of shale and rhyolite greatly increased. These two raw materials are suitable to make elongate blades and stemmed points. It is considered that he hunter-gatherer started to use raw materials selectively depending on the type of tool, especially the newly produced stemmed points. Raw material availability is a crucial factor in determining the trip and direction for hunter-gatherers [@davies2018modeling]. Therefore the adoption of new raw material along with the emergence of new type of tools demonstrates a significant innovation in not just technology but also cognitive system.

@@

1. The new technology and occupation in high elevation

The new technology appeared around 40-35ka in the Korean Peninsula includes stemmed points and following blade industry. The stemmed points brought new paradigms in stone artifact assemblages. For example, projectile points are only a part of composit artifacts and this specific characteristics could contribute to increase efficiency of maintaining the toolkit by easily replacing the damaged part [@kuhn2015artifacts; @cardillo2010some]. Combining with other materials increases utility or functionality of the tool by adding extra weight and providing a shaft, which makes it earlier to grip. Selective use of raw material might impact on hunter-gathers’ mobility and occupation pattern depending on availability of the specific raw materials such as shale and rhyolite. In addition, this research shows that carrying the points enabled hunter-gatherers to occupy higher elevation compared to the sites with no stemmed points. Therefore, we can assume that the technology would play an important role to survive in more risky environments.

2. Population change

Interpreted in HBE, we expected there might be close connection between environmental change and the appearance of new technology based on the prediction that the new invention was the result of risk. The research suggests that the environmental pressure could be related to population change. We cannot simply predicate that the increased population around 35ka was due to dispersal of Homo Sapiens since there is no reliable human remains in South Korea. However, the migration of Homo sapiens was not impossible scenario at the same time. In fact, one model explains that the emergence of stemmed points and blade industry came from outside of the Korean peninsula and the new technology was the result of the continuous influx of modern human populations from Siberia while the earlier coarse flake trandition was come from southern China [@Bae_2010]. Since this migration model barely explains about preexisted stone artifact assemblages and people in the Korean Peninsula, it is hard to accept his argument without knowing about the native people. In short, we assume that the stemmed points appeared related to the increased population and it can be related to the modern human dispersal or indigenously increased occupants.

3. Expedient vs curated

To monitor mobility strategies and site occupation pattern, we examined the artifact density and retouch frequency, based on the premise that highly mobile foragers would have light weighted, highly retouched, replaceable, almost completely made tools and leave small amount of artifacts when they move to the new location (curated). One the other hand, less mobile foragers who tend to stay in one place yield more debitage and debris, produce less efficient tools since they can easily get supplied (expedient) [@clark2019landscapes; @kuhn1994formal; @kuhn2014mousterian; @meignen2006middle]. Therefore, the curated technology indicates sites with special or highly mobile groups with stone artifacts that are thoroughly designed and based on a high level of investment of time and effort in artifact manufacture and maintenance. The expedient technology represents base camp or residence with artifacts that tend to be manufactured, used, and abandoned in the short period of time [@vaquero2018searching;@binford1979organization; @torrence1989time]. There appear to be more variety in the type of sites (or group) during the transition period, both expedient or curated. It became more expedient at the end of the transition, which means foragers became less mobile and tended to stay in the base camp. Combined with the result of population dynamics, we can assume that the increased population made each group bigger and let them stay in one place.
